---
title: "Connecting to a database"
author: "Martijn J. Schuemie"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Connecting to a database}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo = FALSE, message = FALSE}
library(DatabaseConnector)
```

# Introduction

This vignette describes how you can use the `DatabaseConnector` package to connect to a database.

`DatabaseConnector` supports these database platforms:

* Microsoft SQL Server
* Oracle
* PostgresSql
* Amazon Redshift
* Google BigQuery
* DataBricks - Spark
* Snowflake

# Obtaining drivers 

Before `DatabaseConnector` can be used to connect to a database, the drivers for your platform need to be downloaded to a location in the local file system, which we'll refer to as the JAR folder.

## The JAR folder

The JAR folder is just a folder in the local file system where the database drivers are stored. It is highly recommended to use the `DATABASECONNECTOR_JAR_FOLDER` environmental variable to point to this folder, which you can for example set using:

```{r eval=FALSE}
Sys.setenv("DATABASECONNECTOR_JAR_FOLDER" = "c:/temp/jdbcDrivers")
```

Even better would be to add this entry to your `.Renviron` file:

```
DATABASECONNECTOR_JAR_FOLDER = 'c:/temp/jdbcDrivers'
```

That way, the environmental variable will be automatically set whenever you start R. A convenient way to edit your `.Renviron` file is by using `usethis`:

```{r eval=FALSE}
install.packages("usethis")
usethis::edit_r_environ()
```

If you don't use the `DATABASECONNECTOR_JAR_FOLDER` environmental variable, you will need to provide the `pathToDriver` argument every time you call the `downloadJdbcDrivers`, `connect`, `dbConnect`, or `createConnectionDetails` functions.


## Obtaining drivers for SQL Server, Oracle, PostgreSQL, PDW, Spark, RedShift, Snowflake, BigQuery, and Synapse

For your convenience these JDBC drivers are hosted on the OHDSI GitHub pages, and can be downloaded using the `downloadJdbcDrivers` function. You'll first need to specify the JAR folder as described in the previous section, for example using

```{r eval=FALSE}
Sys.setenv("DATABASECONNECTOR_JAR_FOLDER" = "c:/temp/jdbcDrivers")
```

And next download the driver. For example, for PostgreSQL:

```{r eval=FALSE}
library(DatabaseConnector)
downloadJdbcDrivers("postgresql")
```
```{r echo=FALSE}
writeLines("DatabaseConnector JDBC drivers downloaded to 'c:/temp/jdbcDrivers'.")
```

Note that if we hadn't specified the `DATABASECONNECTOR_JAR_FOLDER` environmental variable, we would have to specify the `pathToDriver` argument when calling `downloadJdbcDrivers`.


# Creating a connection

To connect to a database a number of details need to be specified, such as the database platform, the location of the server, the user name, password, and path to the driver. We can call the `connect()` function and specify these details directly:

```{r eval=FALSE}
conn <- DBI::dbConnect(DatabaseConnectorDriver(),
                       dbms = "postgresql",
                       server = "localhost/postgres",
                       user = "joe",
                       password = "secret")
```

```{r echo=FALSE}
writeLines("Connecting using PostgreSQL driver")
```

See [this webpage](http://ohdsi.github.io/DatabaseConnector/reference/connect.html) or type `?connect` for information on which details are required for each platform. Note that we did not need to specify the `pathToDriver` argument because we previously already set the `DATABASECONNECTOR_JAR_FOLDER` environmental variable.

Don't forget to close any connection afterwards:

```{r eval=FALSE}
DBI::dbDisconnect(conn)
```  

Instead of providing the server name, it is also possible to provide the JDBC connection string if this is more convenient:

```{r eval=FALSE}
conn <- DBI::dbConnect(dbms = "postgresql",
                       connectionString = "jdbc:postgresql://localhost:5432/postgres",
                       user = "joe",
                       password = "secret")
```

```{r echo=FALSE}
writeLines("Connecting using PostgreSQL driver")
```

# Using Windows Authentication for SQL Server

In some organizations using Microsoft SQL Server and Windows, it is possible to use Windows Authentication to connect to the server, meaning you won't have to provide a user name and password, since your Windows credentials will be used. This will require downloading the SQL Server authentication DLL file, and placing it somewhere on your system path. If you don't have rights to add files to a place on your system path, you can place it anywhere, and set the `PATH_TO_AUTH_DLL` environmental variable, either using the `Sys.setenv()`, or by adding it to your `.Renviron` file. See [this webpage](http://ohdsi.github.io/DatabaseConnector/reference/connect.html#windows-authentication-for-sql-server-1) or type `?connect` for details on where to get the DLL (and what specific version).
